---
title: "Lecture 1.3 Notebook"
date: 2024-01-08
author: Sofia Ingersoll 
---
## Packages
```{r message = FALSE}
library(tidyverse)
```

## Tidy Data

### Dirty Data

Let's make some untidy data to wrangle with.

Remember, a plot typically requires an x and y axis. In this case, what is our x and y? Do we have more than 2 varibles?

#### What makes this data untidy?
- Each row has more than one observation. We need each observation to make up an entire row.

#### What are we recording? 
- Temperature values need to have its own column.
- An Station ID column should also be included as a column because this is also a variable.
```{r}
# untidy data ----
temp_data_wide <- tribble(
  ~date, ~station1, ~station2,  ~station3,
  "2023-10-01", 30.1, 29.8,  31.2,
  "2023-11-01", 28.6, 29.1,  33.4,
  "2023-12-01", 29.9, 28.5,  32.3
)
```

### Making it Tidy!

We are going to use `pivot_longer` from the tidyr package to clean our data. 
```{r}
# any column name including station, pivot
# and put them in a new column using names_to our new station_id column
# finally, move the values from the old station columns into a new column by the name of temp_c
temp_data_long <- temp_data_wide %>% 
  pivot_longer(cols = starts_with('station'),
               names_to = 'station_id',
               values_to = 'temp_c')
```

