---
title: "EDS 240 Lecture 4 Notebook"
author: "Sofia Ingersoll"
date: 2024-01-29
---

```{r}
install.packages("ggalt") # extra coordinate systems, geoms, statistical transformations, scales and fonts for 'ggplot2' (includes `geom_lollipop()`)
install.packages("ggExtra") # {ggplot2} extension2222 which can be used to add marginal histograms/boxplots/density plots to ggplot2 scatterplots
install.packages("ggdensity") # {ggplot2} extension providing more interpretable visualizations of density estimates based on highest density regions (HDRs)
```

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```{R}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)

#..........................import data...........................
jobs <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/jobs_gender.csv")

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                wrangle data                              ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

jobs_clean <- jobs |> 
  
  # add cols (needed for dumbbell plot) ----
mutate(percent_male = 100 - percent_female, # % of females within each industry was already included
       difference_earnings = total_earnings_male - total_earnings_female) |>  # diff in earnings between M & F
  
  # rearrange columns ----
relocate(year, major_category, minor_category, occupation,
         total_workers, workers_male, workers_female,
         percent_male, percent_female,
         total_earnings, total_earnings_male, total_earnings_female, difference_earnings,
         wage_percent_of_male) |> 
  
  # drop rows with missing earning data ----
drop_na(total_earnings_male, total_earnings_female) |> 
  
  # make occupation a factor ----
mutate(occupation = as.factor(occupation)) |> 
  
  # ---- this next step is for creating our dumbbell plots ----

# classify jobs by percentage male or female ----
mutate(group_label = case_when(
  percent_female >= 75 ~ "Occupations that are 75%+ female",
  percent_female >= 45 & percent_female <= 55 ~ "Occupations that are 45-55% female",
  percent_male >= 75 ~ "Occupations that are 75%+ male"
)) 
```

Let’s first explore the top ten occupations with the highest median earnings in 2016 (full-time workers > 16 years old). The heights of both the bars and lollipops represent the total estimated median earnings (total_earnings). Order them in descending order.
```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#                    Bar & lollipop plots to visualize rankings             ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# Bar Plot
# Top 10 highest earning occupations and order by total earnings
jobs_clean |> 
  filter(year == 2016) |> 
  slice_max(order_by = total_earnings, n = 10) |>      # keep top 10 jobs with most `total_earnings`
  ggplot(
    aes(
      x = fct_reorder(occupation, total_earnings),     # reorder this column (occupation) as a factor by the other column (total earnings)
      y = total_earnings,
      
    )
  ) +
  geom_col() +                                         # creates bar graph with x and y axis. makes   bar height equal to total earnings by each group
  geom_text(
    aes(
      label = scales::dollar(total_earnings),                          # adds $ for values outside of scale functions
    ),
    hjust = 1.2,
    color = 'white'
  ) +
  scale_y_continuous(labels = scales::label_currency(accuracy = 1,       # default sets to a $ in front of y axis
                                                     scale = 0.001,    # reduce the scale of variables by 
                                                     suffix = 'k')) +  # include unit suffix after observations
  coord_flip()




# Lollipop Plot

jobs_clean |> 
  filter(year == 2016) |> 
  slice_max(order_by = total_earnings, n = 10) |> 
  ggplot(
    aes(
      x = fct_reorder(occupation, total_earnings),
      y = total_earnings
    )
  ) +
  ggalt::geom_lollipop() +
  geom_text(
    aes(
      label = scales::dollar(total_earnings),                          # adds $ for values outside of scale functions
    ),
    hjust = -0.2,
    color = 'black'
  ) +
  scale_y_continuous(labels = scales::label_currency(accuracy = 1,       # default sets to a $ in front of y axis
                                                     scale = 0.001,    # reduce the scale of variables by 
                                                     suffix = 'k'),
                     limits = c(0, 225000)
  ) +  # include unit suffix after observations
  coord_flip()
```

## Pivot longer our wide data
```{r}
# 2 groups
#test2 <-
jobs_clean |> 
  filter(year == 2016) |> 
  slice_max(order_by = total_earnings, n = 10) |>      # keep top 10 jobs with most `total_earnings`
  pivot_longer(cols = c(total_earnings_male, total_earnings_female),
               names_to = "group",
               values_to = "earnings_by_group") %>% 
  mutate(sex = str_remove(group, pattern = "total_earnings_")) %>% 
  ggplot(
    aes(
      x = fct_reorder(occupation, earnings_by_group),     # reorder this column (occupation) as a factor by the other column (total earnings)
      y = earnings_by_group,
      fill = sex
    )
  ) +
  geom_col(position = position_dodge()) +                                         # creates bar graph with x and y axis. makes   bar height equal to total earnings by each group
  # geom_text(
  #  aes(
  #   label = scales::dollar(total_earnings),                          # adds $ for values outside of scale functions
  #  ),
  # hjust = 1.2,
  #color = 'white'
  #  ) +
  scale_y_continuous(labels = scales::label_currency(accuracy = 1,       # default sets to a $ in front of y axis
                                                     scale = 0.001,    # reduce the scale of variables by 
                                                     suffix = 'k')) +  # include unit suffix after observations
  coord_flip()
```

```{r}
# Lollipop
jobs_clean |> 
  filter(year == 2016) |> 
  slice_max(order_by = total_earnings, n = 10) |>      # keep top 10 jobs with most `total_earnings`
  pivot_longer(cols = c(total_earnings_male, total_earnings_female),
               names_to = "group",
               values_to = "earnings_by_group") %>% 
  mutate(sex = str_remove(group, pattern = "total_earnings_")) %>% 
  ggplot(
    aes(
      x = fct_reorder(occupation, earnings_by_group),     # reorder this column (occupation) as a factor by the other column (total earnings)
      y = earnings_by_group,
      fill = sex
    )
  ) +
  geom_point(aes(col = sex),
             position = position_dodge(width = 0.5)) +
  geom_linerange(
    aes(
      xmin = occupation,
      xmax = occupation,
      ymin = 0,
      ymax = earnings_by_group,
      col = sex
    ),
    position = position_dodge(width = 0.5)
  ) +
  coord_flip()

```

## Dumbell Plot

```{r}
#....guarantee the same random samples each time we run code.....
set.seed(0)

#.........get 10 random jobs that are 75%+ female (2016).........
f75 <- jobs_clean |> 
  filter(year == 2016, 
         group_label == "Occupations that are 75%+ female") |> 
  slice_sample(n = 10)

#..........get 10 random jobs that are 75%+ male (2016)..........
m75 <- jobs_clean |> 
  filter(year == 2016, 
         group_label == "Occupations that are 75%+ male") |> 
  slice_sample(n = 10)

#........get 10 random jobs that are 45-55%+ female (2016).......
f50 <- jobs_clean |> 
  filter(year == 2016, 
         group_label == "Occupations that are 45-55% female") |> 
  slice_sample(n = 10)

#.......combine dfs & relevel factors (for plotting order).......
subset_jobs <- rbind(f75, m75, f50) |> 
  mutate(group_label = fct_relevel(group_label,
                                   "Occupations that are 75%+ female", 
                                   "Occupations that are 45-55% female", 
                                   "Occupations that are 75%+ male")) 

#Create dumbbell plot
# initialize plot (we'll map our aesthetics locally for each geom, below) ----
ggplot(subset_jobs) +
  
  # create dumbbells ----
geom_segment(aes(x = total_earnings_female,                    # start point of x axis
                 xend = total_earnings_male,                   # end point of x axis
                 y = fct_reorder(occupation, total_earnings),  # reorder occupation by avg_salary here
                 yend = occupation)) +                         # set length of y axis
  geom_point(aes(x = total_earnings_male,                    
                 y = occupation),                              # point to represent male total earnings by occupation
             color = "#CD93D8",
             size = 2.5) +
  geom_point(aes(x = total_earnings_female,       
                 y = occupation),                              # point to represent female total earnings by occupation
             color = "#6A1E99",
             size = 2.5) +
  
  # facet wrap by group ----
facet_wrap(~group_label, 
           nrow = 3,
           scales = "free_y") +                          # "free_y" plots only the axis labels that exist in each group
  
  # axis breaks & $ labels ----
scale_x_continuous(labels = scales::label_dollar(scale = 0.001,
                                                 suffix = "k"),
                   breaks = c(25000, 50000, 75000, 100000, 125000)) +
  labs(title = 'Exploring Gender Wage Disparities by Occupation',
       x = "Total earnings ($)",
       y = "Occupation Title"
       )

#Dumbbell plot output
#There are definitely some additional modifications we would need to make before calling this plot done (e.g. adding a title, legend information, theme) – we’ll come back to that in a later lecture.

17 / 20
```









Make space for long x-axis labels
Give those long x-axis labels some breathing room using coord_flip(), which flips cartesian (x,y) coordinates so that the horizontal becomes the vertical and vice versa.
```{r}
jobs_clean |> 
  filter(year == 2016) |> 
  slice_max(order_by = total_earnings, n = 10) |> 
  ggplot(aes(x = occupation, y = total_earnings)) +
  geom_col() +
  coord_flip()


jobs_clean |> 
  filter(year == 2016) |> 
  slice_max(order_by = total_earnings, n = 10) |> 
  ggplot(aes(x = occupation, y = total_earnings)) +
  ggalt::geom_lollipop() +
  coord_flip()


5 / 20

```